Ecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids

### 2024-05-17 qmd/projects/msu/wss/convert.Rmd
updated website for last time today hopefully 7/17/24

```{r convert2}
# create a word document version of this document
# first need to copy and paste all of the content from this 
# document into an .RMD file so that I can convert the .RMD
# into a word document
library(rmarkdown)
content <- readLines("paper1.qmd", warn = FALSE)
writeLines(content, "paper1.rmd")
render("paper1.rmd", 
       output_format = "word_document",
       output_file = paste0("JacksonStrand_Paper 1.docx"))
render("paper1.rmd", output_format = "pdf_document",
       output_file = paste0("JacksonStrand_Paper 1.pdf"))

>>>>

```{r convert2}
# create a word document version of this document
# first need to copy and paste all of the content from this 
# document into an .RMD file so that I can convert the .RMD
# into a word document
library(rmarkdown)
content <- readLines("paper_1.qmd", warn = FALSE)
writeLines(content, "paper1.rmd")
render("paper1.rmd", 
       output_format = "word_document",
       output_file = paste0("JacksonStrand_Paper 1.docx"))
render("paper1.rmd", output_format = "pdf_document",
       output_file = paste0("JacksonStrand_Paper 1.pdf"))

### 2024-05-17 qmd/projects/msu/wss/paper_1.qmd
updated website for last time today hopefully 7/17/24

---
title: "Ecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
authors: "Jackson R. Strand, David K. Weaver"
date: "May 14, 2024"
editor_options: 
  chunk_output_type: console
bibliography: proposal.bib
link-citations: true
biblio-style: apalike
cls: 
  max-names: 1
---

@@

Numerous polyphagous insect herbivores and their natural enemies rely on a variety of resources found across diverse habitat types within their environment. Depending upon the quality of the resource, different habitat types can serve as sources for increased pest or natural enemy populations. Similarly, habitat types can serve as sinks for decreased populations [@Kennedy2000]. Altering the presence of these resources can have impacts on pest or natural enemy populations and can alter the ecology between the examined species. For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorias2023].

Understanding the ways in which landscape habitat diversity impacts the ecological relationship between pest and enemy is crucial for effectively managing pest species in large scale cropping systems. Similarly, understanding how herbivore plant-hosts act as reservoirs for pests or beneficial natural enemies can potentially play a role in stabilizing year-over-year agricultural impact from these pests [@Gordon1944; @Cano2022].

The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Lesieur2016]. Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Criddle1993]. Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1954]. Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011] and stem toppling due to late-season larval girdling [@Holmes1960].

Several parsitoid species attack *C. cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill 1998]. These two host-specific ectoparasitoids are native to North America. Both parasitoids are bivoltine, with the first generation present in Montana from late June to late July, while the second generation is active in August and September [@Somsen1956]. Adult female *B. cephi* and *B. lissogaster* paralyze *C. cinctus* larvae within the host stem, paralyze the host larvae, and depositing 1 (*B. cephi*) to three (*B. lissogaster*) eggs. *B. cephi* and *B. lissogaster* have been shown to provide biological control via irreplaceable mortality of *C. cinctus* populations [@Buteler2015] and can significantly reduce *C. cinctus* damage and impact on crop yields [@Buteler2008; @Rand2020]. While *C. cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*)[@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts. Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Bhandari2020].

Previous research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B. inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C. cinctus* and Bracon spp. parasitoids quite well [@Peirce2022; @Rand2024]. We find investigating *B. inermis* holds the most potential, as it has been shown to be a quality host for *C. cinctus* while also leading to high larval mortality [@Farstad1945]. However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B. inermis* at multiple time points throughout the growing season. Quantificative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B. inermis* along roadways and within coulles.

@@

Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923]. As both *C. cinctus* and its host-specific parasitoids are relatively restricted in their annual mobility [@Weaver2005], environmental extremes can have broad impacts on both local pest and natural enemy populations. As with many predator-prey interactions, the *C. cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession. Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1991].

@@

```{r}
ratio_plot_data %>%
  dplyr::select(year, region, plant_type, ratio) %>%
  group_by(collection, year, region, plant_type) %>%
  pivot_wider(names_from = "plant_type",
              values_from = "ratio")

lcm <- function(a, b) {
  max_num <- max(a, b)
  while (TRUE) {
    if (max_num %% a == 0 && max_num %% b == 0) {
      return(max_num)
    }
    max_num <- max_num + 1
  }
}

lcm(5, 0.310)
0.310*3.22580645
5*3.22580645
```

>>>>

---
title: "Ecological Importance of Smooth Brome in Managing Wheat Stem Sawfly (Hymenoptera: Cephidae) via Its Associated Braconid Parasitoids"
authors: "Jackson R. Strand, David K. Weaver"
date: "May 17, 2024"
editor_options: 
  chunk_output_type: console
bibliography: paper1.bib
link-citations: true
biblio-style: apalike
cls: 
  max-names: 1
---

@@

Numerous polyphagous insect herbivores and their natural enemies rely on a variety of resources found across diverse habitat types within their environment. Depending upon the quality of the resource, different habitat types can serve as sources for increased pest or natural enemy populations. Similarly, habitat types can serve as sinks for decreased populations [@Kennedy2000]. Altering the presence of these resources can have impacts on pest or natural enemy populations and can alter the ecology between the examined species. For example, diversifying the edge cropping of organic Kale (*Brassica oleracea*) using fava beans, fennel, and marigolds results in a diversity in beneficial arthropod communities that help reduce phytophagous herbivory [@DeMorais2023].

Understanding the ways in which landscape habitat diversity impacts the ecological relationship between pest and enemy is crucial for effectively managing pest species in large scale cropping systems. Similarly, understanding how herbivore plant-hosts act as reservoirs for pests or beneficial natural enemies can potentially play a role in stabilizing year-over-year agricultural impact from these pests [@Cano2022].

The wheat stem sawfly (WSS), Hymenoptera *Cephus cinctus* Norton, is a grass-mining sawfly native to North America [@Lesieur2016]. Originally adept at utilizing wild grass hosts, the species became a major pest to cultivated cereals as the practice became more prevalent following European settlement, becoming a notable pest of winter wheat in 1990 [@Morrill1996]. Adult female WSS use their saw-like ovipositor to lay their eggs in the lumen of stems, where the eggs hatch, leading to larval feeding of parenchyma tissue [@Holmes1956]. Economic damage manifests in the form of both kernel weight reduction caused by tissue damage [@Beres2011] and stem toppling due to late-season larval girdling [@Holmes1960].

Several parsitoid species attack *C. cinctus* within wild grass hosts [@Davis1955; @Morrill1998], however, two species, *Bracon cephis* (Gahan) and *Bracon lissogaster* Muesebeck (Hymenoptera: Braconidae), have been observed targeting WSS within cultivated cereals [@Nelson1953; @Morrill1998]. These two host-specific ectoparasitoids are native to North America. Both parasitoids are bivoltine, with the first generation present in Montana from late June to late July, while the second generation is active in August and September [@Somsen1956]. Adult female *B. cephi* and *B. lissogaster* paralyze *C. cinctus* larvae within the host stem, paralyze the host larvae, and depositing 1 (*B. cephi*) to three (*B. lissogaster*) eggs. *B. cephi* and *B. lissogaster* have been shown to provide biological control via irreplaceable mortality of *C. cinctus* populations [@Buteler2015] and can significantly reduce *C. cinctus* damage and impact on crop yields [@Buteler2008; @Rand2020]. While *C. cinctus* has a wide range of viable host plants, including native and non-native grasses and cultivated row-crops such as wheat (*Triticum aestivum L.*) and barley (*Hordeum vulgare*)[@Wallace1966; @Cockrell2017], infestation levels can vary widely across hosts. Previous research has shown preference for host plants with larger stem diameter [@Ainslie1920; @Bhandari2020].

Previous research has shown natural non-cultivated grasses such as *Agropyron cristatum*, *B. inermis*, *Elymus trachycaulus*, and *Elytrigia intermedium* host *C. cinctus* and Bracon spp. parasitoids quite well [@Peirce2021; @Rand2024]. We find investigating *B. inermis* holds the most potential, as it has been shown to be a quality host for *C. cinctus* while also leading to high larval mortality [@Farstad1945]. However, there is a lack of refined data relating to the total infestation, cutting, and parasitization within *B. inermis* at multiple time points throughout the growing season. Quantificative assessment of potential 'reservoire' characteristics and local spatial information could prove useful in fully understanding the ecological impacts of maintaining *B. inermis* along roadways and within coulles.

@@

Montana's golden triangle, an area of central and northern Montana with intense cereal grain production, is prone to drought and burning via wildfire [@Wilson1923]. As both *C. cinctus* and its host-specific parasitoids are relatively restricted in their annual mobility [@Weaver2005], environmental extremes can have broad impacts on both local pest and natural enemy populations. As with many predator-prey interactions, the *C. cinctus* and parasitoid populations fluctuate, with crashes and spikes experienced in close succession. Similar examples can be found in wolves and moose on Isle Royale [@Peterson1999], and Coccinelidae and aphids [@Evans1999].

@@
